The Talking Heads Experiment

### 2015-04-10 chap1/chap1.tex
cumulative changes to preface and chapter 1

For centuries, philosophers, linguists, psychologists and 
neuroscientists have been groping with the amazing capacities
of the mind. By necessity, they have been doing this through
thought experiments or by observing human behaviour and
brain anatomy. Although this has generated a wealth
of insights,\footnote{Attempts are made to bring together the insights from various
disciplines to establish a true `cognitive science'. 
\cite{Osherton:1995} contains an introduction into some of the main 
research trends in this very diverse scientific field. 
See also \cite{Luger:1994}. The work reported in the present book can be 
classified as theoretical
cognitive science because I try to formulate and test
the operational adequacy of possible models for the 
origins of language and meaning but do not claim nor
give any evidence that these models are also valid for
human cognition, just as the study of aerodynamics and 
aircraft design may help to understand how birds can 
fly but is more generic than its biological 
implementations.}
everyone involved in this research must surely 
agree that we are still lacking adequate models, 
particularly for higher order cognitive functions like language
processing, and definitely for understanding how such 
functions may have arisen. 
To discover or test such models, it therefore
remains useful to do experiments with artificial systems.
We can build robots that receive sensory inputs
through a camera or other sensors, give them computational 
power and memory, and empower them 
for action in the world by adding actuators. Given such a 
set-up, we can precisely examine the operational adequacy 
of an hypothesis. For example, if someone proposes
a process for segmenting images, we can test this process 
by capturing streams of images through a camera and 
see whether the process is indeed capable of performing 
segmentation. When someone proposes a process 
for parsing sentences, we 
can implement this process and confront it with a series of 
example sentences to measure success and failure. 
How else can we test the operational adequacy of a proposed
cognitive model, given the enormous complexity involved? 
Of course, building an artificial system in no way proves
that the principles that were used to construct it are valid
for natural systems. But it is an enormously valuable 
source of insight for approaching the extraordinarily 
complex phenomena observed in human cognition.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap1/figs/heads}}
\caption{Two Talking Heads are shown {\bfshape a1} and 
{\bfshape a2} each seeing the scene on the white board from a slightly different viewpoint.}
\label{f:plate1}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap1/figs/head}}
\caption{ A single `Talking Head'. There is a camera oriented towards a
white board. The bottom screen shows what the camera observes. The top right screen shows the result of processing.
A loudspeaker reproduces the utterances of the agent. }
\label{f:plate2}
\end{figure}

The agents' categorizations of
the world and their language is {\itshape not} programmed but
emerges. It is 
constructed and learned by the agents themselves. The more 
interactions they have with other humans the more they 
adopt our concepts and language. Interacting with the Talking 
Heads is a bit like interacting with two year old 
twins; they play  most of the time with each other and
develop their own language in the process, but the more
humans engage in interaction with them, the more the language
resembles existing human languages.\footnote{Often twins develop a private language, particularly 
if they do not interact much with adults. Compared
to other children, twins are usually 6 months behind 
in their language development but that delay and 
also the private languages that go with it disappear
by the age of 8.}

@@

A fourth radical idea concerns the nature and origins of 
grammar. Rather than invoking the need of a highly 
specialised genetically determined language 
organ\footnote{
As strongly argued by Chomsky in various writings. 
See for example \cite{Chomsky:1968}. 
Praeger, New York. Even though Chomsky argues for 
an innate language acquisition device he has expressed
scepticism about evolution by natural selection
as an explanation for the 
origin of this device (and therefore of language). 
A genetic theory of language evolution 
has been suggested by \cite{Pinker:1994}.}, 
I believe that grammar spontaneously arises when 
generic capabilities to categorise reality, store past
events in terms of abstract schemas, remember associations
between events, etc., reach a critical level and are
applied to language itself. These capabilities are 
relevant across many different cognitive domains. 
In order to store linguistic
experiences, human memories spontaneously structure them, thus 
introducing abstract schemas, internal categories, and 
roles that substructures can play in schemas.
These organisational elements then become externalised. 
Categories are marked by enriching
the form of words, schema boundaries are marked by 
imposing patterns on the expression of a schema, 
roles are marked by assigning
them to specific positions in a 
pattern. This externalisation 
increases the reliability in communication because
it reduces ambiguities and supplies additional context. It 
also aids the stable transmission of the language from one
generation to the next because the language learner
gets additional hints to guess the meaning and function of unknown
words and constructions. 
Once such structuring devices are in place, they help 
to increase the expressive power of the language. The 
complexity of possible language interactions can increase, 
and words or word groups which had multiple usages
can become specialised for particular semantic functions.

@@

\section{A bottom-up approach to Artificial Intelligence}

@@

In 1996, a computer program called Deep Blue
defeated the reigning world champion Kasparov in a game of 
chess.\footnote{See: \cite{Newborn:1996}.}
Kasparov was astonished
and depressed, and claimed this was a defeat of humans in 
the 
race against machines. But was he actually beaten by {\itshape artificial}
intelligence? Not really. A team of engineers and scientists
from Carnegie Mellon University and from the IBM Watson 
research center had been working for ten years to program 
vast amounts of chess knowledge, invented by
human experts, into Deep Blue. They had built extremely 
sophisticated dedicated computing hardware to
apply this knowledge at a blinding speed. So, in beating Kasparov, 
other humans were the clever ones, not machines.

@@

The Talking Heads experiment is the culmination of one of
the most exciting scientific and engineering projects I have
ever been involved in. It has required the
creative efforts of a dozen excellent researchers 
over many years. The story started in 
1985. Instead of continuing to design and program intelligence
explicitly based on formalising human cognitive
capacities, as most of my colleagues
in artificial intelligence research 
labs were doing,\footnote{
Some recent overviews of this `classical' approach to 
Artificial Intelligence can be found in: \cite{Nilsson:1998} and 
\cite{Russell:1998}.}
I started to focus on the 
question of how intelligence might
originate and evolve in physical agents as they 
interact autonomously with their environment or with 
other humans, and I encouraged my students
at the Artificial Intelligence Laboratory of the University 
of Brussels (VUB) to experiment in the same direction. 
We initially developed a bottom-up, 
behaviour-oriented approach to sensori-motor intelligence, 
which was also being explored around the same time by 
Rodney Brooks at the MIT Artificial Intelligence 
laboratory.
The behaviour-oriented, bottom-up approach \is{behaviour-based robots} was a 
counter reaction to the symbolic, top-down approach of 
earlier AI research. See \cite{Steels:1995} and \cite{Arkin:1998}. 
We built robots of various sizes and shapes, using simple
electronic circuits, Lego bricks, small motors, rechargeable
batteries, self-made 
sensors, single board computers, and everything else that 
appeared useful (\figref{f:plate3}).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap1/figs/robot}}
\caption{ Example of a {\it Lego vehicle} built by Tim Smithers based on Lego bricks, a sensori-motor processing board, 
and a variety of 
sensors and actuators. We developed these robots in the early nineties for exploring a behaviour-oriented 
approach to robotics.}
\label{f:plate3}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.85\textwidth]{chap1/figs/fish}}
\caption{ `Artificial fish' built by Miles Pebody to explore influence of robot bodies on behaviour. The fish could swim around by wagging 
its tail and avoid obstacles based on infrared sensing.}
\label{f:plate4}
\end{figure}

To investigate the role of the environment in shaping the
evolving sensori-motor capacities of these robots, 
we built various robotic 
ecosystems in which robots could recharge themselves but also
had to work for their living by 
dimming lights that took away energy from the total energy 
flowing in their ecosystem  (See Figure \ref{f:plate6}). Visitors
to our lab could see robots helping each other or 
engaging in fierce competition for
the resources available for survival. In all of this
research, we tried to see how far behaviours would
autonomously evolve, in other words we tried to find 
mechanisms by which the robots would bootstrap themselves 
towards greater sensori-motor complexity. One of the 
main lessons from these experiments was that 
explanations for cognition lie partly outside the brain 
of the individual agent: The environment, the body, the 
sensori-motor apparatus and the behaviour of the other 
agents all partly shaped their capacities and 
further development.\footnote{This approach is also known as {\it situated cognition}. 
\cite{Clancey:1997}, see also: \cite{Varela:1991}.}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap1/figs/ecosystem}}
\caption{ This ecosystem was designed together with David McFarland (Oxford University) to explore 
emergent cooperation and competition. The robots in the form of lego vehicles could recharge themselves in the charging 
station (shown in the right top corner). But they had to ensure that there was enough energy in the charging 
station by dimming a light in black cylinders by pushing against them.}
\label{f:plate6}
\end{figure}

Many fascinating tales can be told about this research but the intelligence being exhibited 
by these autonomous robots hardly seemed worthy of the name. Yes, they learned
by themselves how to avoid obstacles, how to recharge themselves
in a charging station, or how to co-ordinate efforts 
to exploit the resources in the ecosystems we had built for 
them. But critical observers did not see much more than 
rat intelligence and they were right. The original 
goal of reaching human 
cognitive levels, as observable for instance in 
expert problem solving or conversations in natural
language, remained elusive. The 
traditional Artificial Intelligence
approach of explicitly programming symbolic intelligence
still gave far superior performance in tasks requiring
cognition. Clearly, essential theoretical concepts were 
missing for a truly bottom-up approach to succeed.

In the summer of 1995, a clear breakthrough occurred.  
I was working as a visiting researcher in the Sony
Computer Science laboratory in Tokyo, invited by its director
Mario Tokoro. Reflecting on our 
experiments from a distance, two new ideas occurred to me. 
First of all, language may have been the missing key
in the initial experiments. Language may 
be a necessary route by which the human cognitive system
bootstraps itself autonomously, in tight interaction with the
environment and aided by a community of other
language speakers. This suggested that if we wanted
to have emergent
forms of cognitive intelligence, we needed to go the same 
route. Second, the principles and mechanisms that had been
pouring out of the study of complexity had to be
relevant to understanding the origins and evolution of language,
because they provided generic explanations for how complexity 
may emerge. These principles include 
self-organisation, structural coupling, 
selectionism, level formation, 
and many others (\cite{Nicolis:1989}). The field of {\it Artificial Life} brings together researchers
exploring the insights of complex systems with computational
and robotic experiments (see: \cite{Langton:1995}).

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap1/figs/eye}}
\caption{ First prototype of a {\it Talking Head} camera with 
associated electronics built by Tony Belpaeme. The camera is capable to track moving images. }
\label{f:plate7}
\end{figure}

As my research program grew more radical, it became more 
and more difficult to get funding. 
European research and development programs were 
increasingly demanding short-term projects that targeted information
technology products already available on the American or Japanese 
market. Increasingly, my
research proposals were being rejected and running projects were
cut off,
as reviewers could not see direct short-term 
commercial benefits. All this was endangering the further existence
of my Brussels laboratory, whereas I, paradoxically, thought that
our research had never been more promising. 
Fortunately at this critical moment Mario Tokoro
helped me again in a crucial way. He understood what I was
hoping to achieve and ensured secure and stable resources from 
the Sony Corporation. At the end of 1996, I consequently set up a new research
structure in Paris, a spin-off from the Sony Computer Science
Laboratory in Tokyo, where the bulk of the research reported
in this book could be done in almost ideal circumstances. 
The Paris research was complemented with many important 
contributions from my graduate students
at the VUB AI laboratory in Brussels.

@@

In 1999, a golden opportunity presented itself to 
expose our theories and systems to public scrutiny and thus solicit
judgements from a wide range of human observers. Barbara
Vanderlinden and Hans-Ulrich Obrist, two internationally 
renowned young curators, had been put in charge of 
an important art event in the city of Antwerp (Belgium)
and made the brilliant decision to organise a 
confrontation/co-operation between art and 
science.\footnote{The catalog of this event (\cite{Obrist:1999}), gives an idea of 
the other laboratories and the coming together of art and science.}
They invited \is{Laboratorium exhibition}
artists and scientists to set up 
a public laboratory in the city and conduct experiments 
from the viewpoint of their discipline. This is how 
the Talking Heads experiment came to be installed in 
a public space in Antwerp and how thousands of people, 
some more bewildered than others, took part in the first
ever large-scale public experiment in artificial intelligence. 
Everybody was encouraged to interact with the robots and 
to try and understand what was going on. This was not an obvious
thing to do because the Talking Heads
construct their own language and their own conceptualisation of 
the world. Understanding what they are talking about, resembles the work 
of an anthropologist who is studying the language and
conceptualisations of a newly discovered tribe living
secluded in the rainforest.

@@

This book starts with a preview of the 
experiment and a brief illustration of the core ideas (Chapter 2). 
I then cover the 
different tasks step-by-step that speakers and hearers must carry out: 
perception (Chapter 3), conceptualisation (Chapter 4), 
and lexicalisation (Chapter 5). Each 
chapter discusses the 
architectural components with which the Talking Heads 
have been endowed and the kinds of cognitive structures that 
the agents generate in interaction with the environment 
and the other agents. Chapter 6 and 7 then bring all these 
results together and tests the rich complex semiotic dynamics that 
arise when the Talking Heads effectively interact 
with real world environments.

>>>>

For centuries, philosophers, linguists, psychologists and 
neuroscientists have been groping with the amazing capacities
of the mind. By necessity, they have been doing this through
thought experiments or by observing human behaviour and
brain anatomy. Although this has generated a wealth
of insights,\footnote{Attempts are made to bring together the insights from various
disciplines to establish a true ``cognitive science''. 
\cite{Osherton:1995} contains an introduction into some of the main 
research trends in this very diverse scientific field. 
See also \cite{Luger:1994}. The work reported in the present book can be 
classified as theoretical
cognitive science because I try to formulate and test
the operational adequacy of possible models for the 
origins of language and meaning but do not claim nor
give any evidence that these models are also valid for
human cognition, just as the study of aerodynamics and 
aircraft design may help to understand how birds can 
fly but is more generic than its biological 
implementations.}
everyone involved in this research must surely 
agree that we are still lacking adequate models, 
particularly for higher order cognitive functions like language
processing, and definitely for understanding how such 
functions may have arisen. 
To discover or test such models, it therefore
remains useful to do experiments with artificial systems.
We can build robots that receive sensory inputs
through a camera or other sensors, give them computational 
power and memory, and empower them 
for action in the world by adding actuators. Given such a 
set-up, we can precisely examine the operational adequacy 
of an hypothesis. For example, if someone proposes
a process for segmenting images, we can test this process 
by capturing streams of images through a camera and 
see whether the process is indeed capable of performing 
segmentation. When someone proposes a process 
for parsing sentences, we 
can implement this process and confront it with a series of 
example sentences to measure success and failure. 
How else can we test the operational adequacy of a proposed
cognitive model, given the enormous complexity involved? 
Of course, building an artificial system in no way proves
that the principles that were used to construct it are valid
for natural systems. But it is an enormously valuable 
source of insight for approaching the extraordinarily 
complex phenomena observed in human cognition.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{./chap1/figs/heads.pdf}}
\caption{Two Talking Heads are shown {\bfshape a1} and 
{\bfshape a2} each seeing the scene on the white board from a slightly different viewpoint.}
\label{f:plate1}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{./chap1/figs/head.pdf}}
\caption{ A single `Talking Head'. There is a camera oriented towards a
white board. The bottom screen shows what the camera observes. The top right screen shows the result of processing.
A loudspeaker reproduces the utterances of the agent. }
\label{f:plate2}
\end{figure}

The agents' categorisations of
the world and their language is {\itshape not} programmed but
emerges. It is 
constructed and learned by the agents themselves. The more 
interactions they have with other humans the more they 
adopt our concepts and language. Interacting with the Talking 
Heads is a bit like interacting with two year old 
twins; they play  most of the time with each other and
develop their own language in the process, but the more
humans engage in interaction with them, the more the language
resembles existing human languages.\footnote{Often twins develop a private language, particularly 
if they do not interact much with adults. Compared
to other children, twins are usually 6 months behind 
in their language development but that delay and 
also the private languages that go with it disappear
by the age of 8.}

@@

A fourth radical idea concerns the nature and origins of 
grammar. Rather than invoking the need of a highly 
specialised genetically determined language 
organ,\footnote{
As strongly argued by Chomsky in various writings. 
See for example \cite{Chomsky:1968}. 
Praeger, New York. Even though Chomsky argues for 
an innate language acquisition device he has expressed
scepticism about evolution by natural selection
as an explanation for the 
origin of this device (and therefore of language). 
A genetic theory of language evolution 
has been suggested by \cite{Pinker:1994}.} 
I believe that grammar spontaneously arises when 
generic capabilities to categorise reality, store past
events in terms of abstract schemas, remember associations
between events, etc., reach a critical level and are
applied to language itself. These capabilities are 
relevant across many different cognitive domains. 
In order to store linguistic
experiences, human memories spontaneously structure them, thus 
introducing abstract schemas, internal categories, and 
roles that substructures can play in schemas.
These organisational elements then become externalised. 
Categories are marked by enriching
the form of words, schema boundaries are marked by 
imposing patterns on the expression of a schema, 
roles are marked by assigning
them to specific positions in a 
pattern. This externalisation 
increases the reliability in communication because
it reduces ambiguities and supplies additional context. It 
also aids the stable transmission of the language from one
generation to the next because the language learner
gets additional hints to guess the meaning and function of unknown
words and constructions. 
Once such structuring devices are in place, they help 
to increase the expressive power of the language. The 
complexity of possible language interactions can increase, 
and words or word groups which had multiple usages
can become specialised for particular semantic functions.

@@

\section{A bottom-up approach to artificial intelligence}

@@

In 1996, a computer program called Deep Blue
defeated the reigning world champion Kasparov in a game of 
chess.\footnote{See: \cite{Newborn:1996}.}
Kasparov was astonished
and depressed, and claimed this was a defeat of humans in 
the 
race against machines. But was he actually beaten by {\itshape artificial}
intelligence? Not really. A team of engineers and scientists
from Carnegie Mellon University and from the IBM Watson 
Research Center had been working for ten years to program 
vast amounts of chess knowledge, invented by
human experts, into Deep Blue. They had built extremely 
sophisticated dedicated computing hardware to
apply this knowledge at a blinding speed. So, in beating Kasparov, 
other humans were the clever ones, not machines.

@@

The Talking Heads experiment is the culmination of one of
the most exciting scientific and engineering projects I have
ever been involved in. It has required the
creative efforts of a dozen excellent researchers 
over many years. The story started in 
1985. Instead of continuing to design and program intelligence
explicitly based on formalising human cognitive
capacities, as most of my colleagues
in artificial intelligence research 
labs were doing,\footnote{
Some recent overviews of this `classical' approach to 
artificial intelligence can be found in: \cite{Nilsson:1998} and 
\cite{Russell:1998}.}
I started to focus on the 
question of how intelligence might
originate and evolve in physical agents as they 
interact autonomously with their environment or with 
other humans, and I encouraged my students
at the Artificial Intelligence Laboratory of the University 
of Brussels (VUB) to experiment in the same direction. 
We initially developed a bottom-up, 
behaviour-oriented approach to sensori-motor intelligence, 
which was also being explored around the same time by 
Rodney Brooks at the MIT Artificial Intelligence 
Laboratory.
The behaviour-oriented, bottom-up approach \is{behaviour-based robots} was a 
counter reaction to the symbolic, top-down approach of 
earlier AI research. See \cite{Steels:1995} and \cite{Arkin:1998}. 
We built robots of various sizes and shapes, using simple
electronic circuits, Lego bricks, small motors, rechargeable
batteries, self-made 
sensors, single board computers, and everything else that 
appeared useful (\figref{f:plate3}).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap1/figs/robot.pdf}}
\caption{ Example of a {\it Lego vehicle} built by Tim Smithers based on Lego bricks, a sensori-motor processing board, 
and a variety of 
sensors and actuators. We developed these robots in the early nineties for exploring a behaviour-oriented 
approach to robotics.}
\label{f:plate3}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.85\textwidth]{chap1/figs/fish.pdf}}
\caption{ `Artificial fish' built by Miles Pebody to explore influence of robot bodies on behaviour. The fish could swim around by wagging 
its tail and avoid obstacles based on infrared sensing.}
\label{f:plate4}
\end{figure}

To investigate the role of the environment in shaping the
evolving sensori-motor capacities of these robots, 
we built various robotic 
ecosystems in which robots could recharge themselves but also
had to work for their living by 
dimming lights that took away energy from the total energy 
flowing in their ecosystem (see \figref{f:plate6}). Visitors
to our lab could see robots helping each other or 
engaging in fierce competition for
the resources available for survival. In all of this
research, we tried to see how far behaviours would
autonomously evolve, in other words we tried to find 
mechanisms by which the robots would bootstrap themselves 
towards greater sensori-motor complexity. One of the 
main lessons from these experiments was that 
explanations for cognition lie partly outside the brain 
of the individual agent: The environment, the body, the 
sensori-motor apparatus and the behaviour of the other 
agents all partly shaped their capacities and 
further development.\footnote{This approach is also known as {\it situated cognition}. 
\cite{Clancey:1997}, see also: \cite{Varela:1991}.}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap1/figs/ecosystem.pdf}}
\caption{ This ecosystem was designed together with David McFarland (Oxford University) to explore 
emergent cooperation and competition. The robots in the form of lego vehicles could recharge themselves in the charging 
station (shown in the right top corner). But they had to ensure that there was enough energy in the charging 
station by dimming a light in black cylinders by pushing against them.}
\label{f:plate6}
\end{figure}

Many fascinating tales can be told about this research but the intelligence being exhibited 
by these autonomous robots hardly seemed worthy of the name. Yes, they learned
by themselves how to avoid obstacles, how to recharge themselves
in a charging station, or how to co-ordinate efforts 
to exploit the resources in the ecosystems we had built for 
them. But critical observers did not see much more than 
rat intelligence and they were right. The original 
goal of reaching human 
cognitive levels, as observable for instance in 
expert problem solving or conversations in natural
language, remained elusive. The 
traditional artificial intelligence
approach of explicitly programming symbolic intelligence
still gave far superior performance in tasks requiring
cognition. Clearly, essential theoretical concepts were 
missing for a truly bottom-up approach to succeed.

In the summer of 1995, a clear breakthrough occurred.  
I was working as a visiting researcher in the Sony
Computer Science Laboratory in Tokyo, invited by its director
Mario Tokoro. Reflecting on our 
experiments from a distance, two new ideas occurred to me. 
First of all, language may have been the missing key
in the initial experiments. Language may 
be a necessary route by which the human cognitive system
bootstraps itself autonomously, in tight interaction with the
environment and aided by a community of other
language speakers. This suggested that if we wanted
to have emergent
forms of cognitive intelligence, we needed to go the same 
route. Second, the principles and mechanisms that had been
pouring out of the study of complexity had to be
relevant to understanding the origins and evolution of language,
because they provided generic explanations for how complexity 
may emerge. These principles include 
self-organisation, structural coupling, 
selectionism, level formation, 
and many others (\citealt{Nicolis:1989}). The field of {\it artificial life} brings together researchers
exploring the insights of complex systems with computational
and robotic experiments (see: \citealt{Langton:1995}).

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap1/figs/eye.pdf}}
\caption{ First prototype of a {\it Talking Head} camera with 
associated electronics built by Tony Belpaeme. The camera is capable to track moving images. }
\label{f:plate7}
\end{figure}

As my research programme grew more radical, it became more 
and more difficult to get funding. 
European research and development programmes were 
increasingly demanding short-term projects that targeted information
technology products already available on the American or Japanese 
market. Increasingly, my
research proposals were being rejected and running projects were
cut off,
as reviewers could not see direct short-term 
commercial benefits. All this was endangering the further existence
of my Brussels laboratory, whereas I, paradoxically, thought that
our research had never been more promising. 
Fortunately at this critical moment Mario Tokoro
helped me again in a crucial way. He understood what I was
hoping to achieve and ensured secure and stable resources from 
the Sony Corporation. At the end of 1996, I consequently set up a new research
structure in Paris, a spin-off from the Sony Computer Science
Laboratory in Tokyo, where the bulk of the research reported
in this book could be done in almost ideal circumstances. 
The Paris research was complemented with many important 
contributions from my graduate students
at the VUB AI Laboratory in Brussels.

@@

In 1999, a golden opportunity presented itself to 
expose our theories and systems to public scrutiny and thus solicit
judgements from a wide range of human observers. Barbara
Vanderlinden and Hans-Ulrich Obrist, two internationally 
renowned young curators, had been put in charge of 
an important art event in the city of Antwerp (Belgium)
and made the brilliant decision to organise a 
confrontation/co-operation between art and 
science.\footnote{The catalogue of this event (\citealt{Obrist:1999}) gives an idea of 
the other laboratories and the coming together of art and science.}
They invited \is{Laboratorium exhibition}
artists and scientists to set up 
a public laboratory in the city and conduct experiments 
from the viewpoint of their discipline. This is how 
the Talking Heads experiment came to be installed in 
a public space in Antwerp and how thousands of people, 
some more bewildered than others, took part in the first
ever large-scale public experiment in artificial intelligence. 
Everybody was encouraged to interact with the robots and 
to try and understand what was going on. This was not an obvious
thing to do because the Talking Heads
construct their own language and their own conceptualisation of 
the world. Understanding what they are talking about resembles the work 
of an anthropologist who is studying the language and
conceptualisations of a newly discovered tribe living
secluded in the rainforest.

@@

This book starts with a preview of the 
experiment and a brief illustration of the core ideas (\chapref{chap:2}). 
I then cover the 
different tasks step-by-step that speakers and hearers must carry out: 
perception (\chapref{chap:3}), conceptualisation (\chapref{chap:4}), 
and lexicalisation (\chapref{chap:5}). Each 
chapter discusses the 
architectural components with which the Talking Heads 
have been endowed and the kinds of cognitive structures that 
the agents generate in interaction with the environment 
and the other agents. \chapref{chap:6} and \chapref{chap:7} then bring all these 
results together and tests the rich complex semiotic dynamics that 
arise when the Talking Heads effectively interact 
with real world environments.

### 2015-04-10 chap8/chap8.tex
cumulative changes to preface and chapter 1

Numerous talks were given and various papers published in scientific fora. There were also various talks within 
the art context\footnote{Examples are a 'gallery talk' as Salon 3 at the Elephant and Castle Centre
in London on 2 december 1998, organised by Hans Ulrich Obrist and Molly Nesbit, and a presentation together with 
Hans-Ulrich Obrist and Rem Koolhaas in Antwerp to launch the Laboratorium book on 3 october 2001.}
The experiment received wide coverage in 
the media thanks to its public exposure as part of a major exhibition. The Sued-Deutsche Zeitung called it 
`Angels with Internet wings'. All this lead to invitations to show the experiment
also in other venues, as discussed in the next chapter. 
Regretably, the challenge just to keep the experiment in the air with the available human resources
and the pressure for going on with new experiments prevented us from doing more adequate data gathering 
and analysis. Nevertheless some analyses appeared, particularly as carried out by Frederic Kaplan\cite{Kaplan:2001}.

>>>>

Numerous talks were given and various papers published in scientific fora. There were also various talks within 
the art context\footnote{Examples are a 'gallery talk' as Salon 3 at the Elephant and Castle Centre
in London on 2 december 1998, organised by Hans Ulrich Obrist and Molly Nesbit, and a presentation together with 
Hans-Ulrich Obrist and Rem Koolhaas in Antwerp to launch the Laboratorium book on 3 October 2001.}
The experiment received wide coverage in 
the media thanks to its public exposure as part of a major exhibition. The Sued-Deutsche Zeitung called it 
`Angels with Internet wings'. All this lead to invitations to show the experiment
also in other venues, as discussed in the next chapter. 
Regretably, the challenge just to keep the experiment in the air with the available human resources
and the pressure for going on with new experiments prevented us from doing more adequate data gathering 
and analysis. Nevertheless some analyses appeared, particularly as carried out by Frederic Kaplan\cite{Kaplan:2001}.
